---
title: "Boss Analysis"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Boss Analysis}
  %\VignetteEncoding{UTF-8}
---

```{r warning = FALSE, message = FALSE}
rm(list = ls())
seed <- 1
set.seed(seed)

require(kumu)
require(stringi)
require(data.table)
require(jsonlite)
require(igraph)
require(visNetwork)
```

# Introduction

This Notebook continues the analysis of Null Variables, utilizing the derived NumEdges1PNEF, i.e. the number of edges on the bootstrapped causal graph using only the variables of interest after applying the 1PNEF threshold. Our goal in this Notebook is to perform no bootstrap, but experiment with 1 RUN and different penalty scores until we obtain a similar number of edges as NumEdges1PNEF of the `random_causality_threshold.Rmd` Notbook, which for Camel 1 was 92 edges.

# Dataset 

First we define the input dataset for Tetrad, and subsequently the output folder path.

```{r}
dt_path <- "~/Downloads/camel_1_6-flaws_smells_vs_outcome_metrics_no_filepath_deleted_columns.csv"
output_folder_path <- "~/projects/kumu_data/boss_causal_discovery/causal_output"
filename <- "causal_graph_camel_1_6"
filepath <- stringi::stri_c(file.path(output_folder_path,filename),"_graph.json")

data_flags <- data_io(dataset_path = dt_path,
                         data_type = "continuous",
                         column_delimiter = "comma",
                         output_folder_path = output_folder_path,
                         filename = filename,
                         is_json_output = TRUE)
```

# Knowledge File

```{r}
#knowledge_file_path <- "~/Downloads/knowledge_2.txt"
knowledge_file_path <- "~/projects/kumu_data/mike_knowledge_box.txt"
knowledge_flags <- knowledge_file_path(knowledge_file_path)
```

# Causal Algorithm

Next, we define the causal algorithm to be used in Tetrad, in this case BOSS. Note contrary to the Notebook where we derived the 1PNEF, we do not bootstrap here or conduct multiple restarts. Rather, we only run the algorithm once, carying the penalty discount until 

```{r}
algorithm_flags <- algorithm_boss(num_starts = 1,
                                  time_lag = 0,
                                  allow_internal_randomness = FALSE,
                                  use_bes = FALSE)
```

# Score 

We subsequently choose the score metric, in this case Sem BIC. We begin with Penalty Discount 1. 

```{r}
score_flags <- score_sem_bic(penalty_discount = 0.06, 
                             sem_bic_rule = 1, 
                             sem_bic_structure_prior = 0, 
                             precompute_covariances = TRUE)
```

# Bootstrapping

Lastly, we choose the bootstrapping criteria. Observe here the parameter is `number_resampling=0`, to indicate no bootstrap is being used.

```{r}
bootstrapping_flags <- bootstrapping(number_resampling=0,
                                     percent_resample_size = 100,
                                     seed = 32, 
                                     add_original_dataset = TRUE,
                                     resampling_with_replacement = TRUE,
                                     resampling_ensemble = 1,
                                     save_bootstrap_graphs = FALSE)

```

# Tetrad Command

```{r results='hide'}
tetrad_path <- "~/Downloads/causal-cmd-1.10.0-jar-with-dependencies.jar"
tetrad(tetrad_cmd_path = tetrad_path,
       data_flags = data_flags, 
   #    knowledge_flags = knowledge_flags,
       algorithm_flags = algorithm_flags,
       score_flags = score_flags,
       bootstrapping_flags = bootstrapping_flags)
```

# Debug

```{r}
stringi::stri_c(c(tetrad_path,data_flags,knowledge_flags,algorithm_flags,score_flags,bootstrapping_flags),collapse = " ")
```
# Read Graph 

```{r}
graph <- parse_graph(filepath)
graph[["nodes"]]
```

```{r}
graph[["edgeset"]]
```
In this case, we expect this table to be empty, as if only 1 run is executed, then the edgeset table contains the only edge the type probabilities table would have. 

```{r}
graph[["edge_type_probabilities"]]
```

Of interest here, the obtained number of edges is:

```{r}
nrow(graph[["edgeset"]])
```

We can then adjust the penalty discount to be within at most 5% off the NumEdges1PNEF. In a future notebook, we can automate the change of PD over a range to generate a line plot describing how the number of rows change as a function of PD. This should simplify the trial and error step. 

# Graph Visualization

```{r}
nodes <- data.table::copy(graph[["nodes"]])
colnames(nodes) <- "node"
```

For the edgelist, it is possible an edge is "tail tail", meaning no causal direction could be determined. Since graph visualizations normally do not allow the display of mixed graph directions, and to facilitate readability, we will color the tail-tail edges in red, and the tail-arrow in black. 

```{r}
#edges <- graph[["edgeset"]][,.(from=node1_name,to=node2_name,value=probability,weight=probability,label=probability)]

edges <- graph[["edgeset"]]
edges$color <- "black"
edges[endpoint1 == "TAIL" & endpoint2 == "TAIL"]$color <- "red"
edges <- edges[,.(from=node1_name,to=node2_name,color=color)]
```


Note the probabilities in this case are 0, since we do not perform multiple runs or bootstrap.

```{r}
g_viz <- igraph::graph_from_data_frame(d=edges, 
                      directed = TRUE, 
                      vertices = nodes)

g_viz <- visIgraph(g_viz,
          randomSeed = 1)#,
          #layout = "layout_with_dh")
#vis_graph <- toVisNetworkData(graph)
#visNetwork(nodes = vis_graph$nodes, edges = vis_graph$edges,randomSeed = 1,
#           height = "600px", width = "100%") %>% 
g_viz %>% visOptions(highlightNearest = TRUE) %>% visInteraction(navigationButtons = TRUE)#  %>% 
  #visHierarchicalLayout()
  #visInteraction(navigationButtons = TRUE,keyboard = TRUE, tooltipDelay = 0 ) 
```

