---
title: "Boss Analysis"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Boss Analysis}
  %\VignetteEncoding{UTF-8}
---

# Introduction

```{r warning = FALSE, message = FALSE}
rm(list = ls())
seed <- 1
set.seed(seed)

require(kumu)
require(stringi)
require(data.table)
```

# Dataset 

First we define the input dataset for Tetrad, and subsequently the output folder path.

```{r}
dt_path <- "~/Downloads/camel_1_6-flaws_smells_vs_outcome_metrics_no_filepath_deleted_columns.csv"
output_folder_path <- "~/Downloads/causal_output/"

data_flags <- data_io(dataset_path = dt_path,
                         data_type = "continuous",
                         column_delimiter = "comma",
                         output_folder_path = output_folder_path,
                         is_json_output = TRUE)
```

# Knowledge File

```{r}
knowledge_file_path <- "~/Downloads/knowledge_2.txt"
knowledge_flags <- knowledge_file_path(knowledge_file_path)
```

# Causal Algorithm

Next, we define the causal algorithm to be used in Tetrad, in this case BOSS.

```{r}
algorithm_flags <- algorithm_boss(num_starts = 1,
                                  time_lag = 0,
                                  allow_internal_randomness = FALSE,
                                  use_bes = FALSE)
```

# Score 

We subsequently choose the score metric, in this case Sem BIC.

```{r}
score_flags <- score_sem_bic(penalty_discount = 1, 
                             sem_bic_rule = 1, 
                             sem_bic_structure_prior = 0, 
                             precompute_covariances = TRUE)
```

# Bootstrapping

Lastly, we choose the bootstrapping criteria.

```{r}
bootstrapping_flags <- bootstrapping(number_resampling=0,
                                     percent_resample_size = 100,
                                     seed = 32, 
                                     add_original_dataset = TRUE,
                                     resampling_with_replacement = TRUE,
                                     save_bootstrap_graphs = FALSE)

```

# Tetrad Command

```{r}
tetrad_path <- "~/Downloads/causal-cmd-1.10.0-jar-with-dependencies.jar"
tetrad(tetrad_cmd_path = tetrad_path,
       data_flags = data_flags, 
   #    knowledge_flags = knowledge_flags,
       algorithm_flags = algorithm_flags,
       score_flags = score_flags,
       bootstrapping_flags = bootstrapping_flags)
```

# Debug

```{r}
stringi::stri_c(c(tetrad_path,data_flags,knowledge_flags,algorithm_flags,score_flags,bootstrapping_flags),collapse = " ")
```

