---
title: "Null Variables"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Null Variables}
  %\VignetteEncoding{UTF-8}
---

# Introduction

First, we load the necessary packages. 

```{r warning = FALSE, message = FALSE}
rm(list = ls())
seed <- 1
set.seed(seed)

require(stringi)
require(data.table)
```

Next, we load the dataset. This dataset can be generated using GitHub's Sailuh/Kaiaulu [causal_flaws.Rmd](https://github.com/sailuh/kaiaulu/blob/master/vignettes/causal_flaws.Rmd) notebook.

```{r}
dt <- fread("~/Desktop/kaiaulu/analysis/dv8/tse_cassandra/tse_cassandra-flaws_smells_vs_outcome_metrics.csv")
```

# Add Null Features

We generate a replica column prefixed by `nv-`. The nv- columns have their values shufled across the rows, hence the null (random) naming to them.

```{r}
nv_dt <- copy(dt[,2:ncol(dt),with=FALSE])
colnames(nv_dt) <- stringi::stri_c("nv-",colnames(nv_dt))
nv_dt <- apply(nv_dt,2,sample)
nv_dt <- cbind(dt,nv_dt)
head(nv_dt)
```

To save, uncomment the following code line and execute where you wish to save it:

```{r}
#fwrite(nv_dt,"~/Downloads/project_nv_dt.csv")
```


